---
title: "Analyzing and Comparing Income Inequality Trends in the Canada: Impacts of the Great Recession and COVID-19 Pandemic"
subtitle: "Insights from 2002 to 2023"
author: Dong Jun Yoon
thanks: "Code and data are available at: https://github.com/DJY1231/Exploring_Income_Inequality_During_Covid_19.git."
date: today
date-format: long
abstract: "This paper investigates income inequality in Canada from 2002 to 2023 using data from the World Inequality Database (WID). Throughout the paper, we will explore trends in income inequality, examining whether it is increasing or decreasing over time, and how the Great Recession of 2007-2009 and the Covid-19 of 2020-2023 affected income inequality. We will utilize data analysis and regression modeling to investigate these trends. Investigating these trends is significant, as it could assist policymakers or the government in developing strategies that could decrease the inequalities occurring within the country. As a result, this report may be used to help promote social and economic equity in Canada."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```


# Introduction
Income inequality is a critical issue both in Canada and all around the world. Income inequality refers to the difference in how income is distributed among the population, highlighting significant disparities in financial well-being. Unlike wealth inequality, which considers the value of assets such as homes, stocks, and other holdings, income inequality specifically examines earnings from sources like wages, salaries, savings interest, stock dividends, rental income, and profits from the sale of assets. This distinction is essential for understanding the multifaceted nature of economic disparities. Gaining a deeper understanding of income inequality can equip policymakers and researchers with valuable insights to develop strategies aimed at addressing this pervasive problem.

This paper focuses on the dynamics of income inequality in Canada, especially focusing on the period from 2002 to 2023. Using the dataset provided by World Inequality Database (WID), the paper will fully investigate income trends across the social class, top 1%, top 10% and bottom 50% of income inequality. Moreover, will be considering variables such as year, social class, and race to find out the correlation between these factors. 

To efficiently investigate remaining income gap between individual firms and social classes who are living in Canada, this paper will focus on exploring datasets and multiple linear regression modeling to find out the full details (trends, patterns) about income inequality that is occurring in Canada. 

The paper is structured as following: Section 2 introduces the data used for analysis and findings, including the variables of interest. Section 3 introduces the linear regression model to examine the correlation between time (year) and social class to investigate income inequality in Canada. Section 4 introduces the full analysis of the table which will include the summary. Section 5 may discuss about the findings that we investigated and also talk about the weaknesses of this paper and its next steps.


# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 



# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


